Infering Fine-Scale Animal Behaviors using Hidden Markov Models

### 2020-06-05 Report/Sections/carHMM.tex
Update simulation study
Many Updates

\section{Autocorrelation within Fine Scale Behaviors}

One of the key assumption of both HMMs and HHMM is \textit{conditional indpendence} between observations. In particular, given the state $X_t$ or $X^*_{s,t}$, $Y_t$ or $Y^*_{s,t}$ (respectively) is assumed to be independent from all other observations. Therefore,traditional HMMs can fail is when the observations $Y$ exhibit significant auto-correlation. Unfortunately this is often the case for fine-scale animal behaviors. Examples include fluking in marine mammals in Vancouver, BC (see the results section) and swiming behavior in horn sharks in southern California \cite{Adam:2019}.

One way to deal with autocorrelation in fine-scale behavioral processes is to use a state-switching continuous model such as the one introduced by Michelot et al \cite{Michelot:2019}, which models the movement of an animal as an Ornstein-Uhlenbeck process with parameters that depend upon the underlying state of the animal. Continuous time models are advantagous because of their flexibiliy: they can be built up from aribitrarily complex stochastic differential equations and they allow for uneven step lengths in the observations sequence $\mathbf{Y}$. However, most continuous time models require MCMC algorithms to perform inference and as a result are not easily incorporated into the HHMM structure.

Another option is to use the CarHMM, or \textit{conditionally auto-regressive hidden Markov model}, introduced by Lawler et al \cite{Lawler:2019}, in which autocorrelation is explicity modeled into the emission distributions of the HMM while maintaining the strucutre needed to run the forward algorithm for fast direct likelihood maximization. In particular, the CarHMM introduces autocorrleation into the HMM by assuming that an observation $Y_t$ has mean $(1-\phi_{x_t}) \cdot \mu_{x_t} + \phi_{x_t} \cdot y_{t-1}$, instead of $\mu_{x_t}$ in the case of a classic HMM. Note that the autocorrelation term $\phi_{x_t}$ depends upon the behavioral state of the animal. This model easily fits into the HHMM structure, but initially it seems to lack the flexibility and natural interpretation of continuous-time models.

\subsection{Equivalency of CarHMM and one-dimensional Ornstein-Uhlenbeck  state-switching process}

In fact, under certain conditions, the CarHMM and the state-switching continuous model are equivalent. In particular, the conditions are that (1) the underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process, and (2) the emission distributions of the CarHMM must be gaussian.

This allows the theoretically grounded continuous-time state-switching model to be used in the computational convieneint HMM (and therefore HHMM) framework. In addition, it gives new intpretation to the learned parameters of the CarHMM in the context of an Ornstein-Uhlenbeck process.

A one-dimensional state-switching Ornstein-Uhlenbeck process $y^*$ is the solution to the following stochastic differential equation:
%
$$dy^*_t = \beta_{x^*_t}(\gamma_{x^*_t} - y^*_t)dt + \omega_{x^*_t} dW_t$$
%
where $x^*_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x^*_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x^*_t}$ is the long-term mean value of the process, $\omega_{x^*_t}$ is related to short-term variance, and $W$ is a Wiener process. As before, $x^*_t$ is described by an unobserved Markov process. The solution to this equation is known to be the following \cite{Michelot:2019}:
\begin{align*}
	y^*_{t+\delta} \sim \mathcal{N}\left((1-e^{-\beta_{x^*_t}\delta})\gamma_{x^*_t} + e^{-\beta_{x^*_t}\delta} y^*_t,\quad \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})\right)
\end{align*}
Now, suppose that $\delta$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
\begin{align*}
	\mu_{x^*_t} = \gamma_{x^*_t}, \qquad \phi_{x^*_t} = e^{-\beta_{x^*_t}\delta}, \qquad \sigma^2_{x^*_t} = \frac{\omega_{x^*_t}^2}{2\beta_{x^*_t}} (1-e^{-2\beta_{x^*_t}\delta})
\end{align*}
Then, we have the following:
\begin{align*}
	y^*_{t+\delta} \sim \mathcal{N}\left((1-\phi_{x^*_t})\mu_{x^*_t} + \phi_{x^*_t} y^*_t,\quad \sigma_{x^*_t}^2 \right)
\end{align*}
%
If $\delta$ is fixed and $x^*_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities. Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again. This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.

>>>>

\section{Results regarding CarHMMs and the Ornstein-Uhlenbeck process}

\subsection{Equivalence of CarHMM and OU process}

Once the CarHMM is fit by maximizing the likelihood in equation (\ref{CarHMM_likelihood}), the emission distributions can be interpretted as the solution of a state-switching Ornstein-Uhlenbeck process similar to the one introduced by Michelot et al \cite{Michelot:2019}. In particular, the following two conditions must be met:
\begin{enumerate}
	\item The underlying behavioral state of the continuous-time model must follow a Markov chain rather than a Markov process.
	\item The emission distributions of the CarHMM must be normal.
\end{enumerate}
If both conditions (1) and (2) are met, then the CarHMM is equivalent to a state-switching Ornstein-Uhlenbeck process. This gives new intpretation to the learned parameters of the CarHMM in the context of continuous-time model.

\subsubsection{Proof}

A one-dimensional state-switching Ornstein-Uhlenbeck process $y$ is the solution to the following stochastic differential equation:
%
$$dy_t = \beta_{x_t}(\gamma_{x_t} - y_t)dt + \omega_{x_t} dW_t$$
%
where $x_t$ is the fine-scale behavior of the animal at time $t$, $\beta_{x_t}$ relates to rate at which the process returns to its mean value, $\gamma_{x_t}$ is the long-term mean value of the process, $\omega_{x_t}$ is related to short-term variance, and $W$ is a Wiener process. $x_t$ is described by an unobserved Markov process. If the behavioral state $x_t$ does not change between observations (i.e. $x_t$ follows a Markov chain), the solution to this equation is known to be the following \cite{Michelot:2019}:
\begin{align*}
y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-e^{-\beta_{x_t}\Delta t})\gamma_{x_t} + e^{-\beta_{x_t}\Delta t} y_t,\quad \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})\right)
\end{align*}
Now, suppose that $\Delta t$ is constant for all observations, as is the case for hidden Markov models. In addition, introduce the following transformations:
\begin{align}
\mu_{x_t} = \gamma_{x_t}, \qquad \phi_{x_t} = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t} = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})
\label{CarHMM_to_OU}
\end{align}
Then, we have the following:
\begin{align*}
y_{t+\Delta t} | x_{t:t+\Delta t} \sim \mathcal{N}\left((1-\phi_{x_t})\mu_{x_t} + \phi_{x_t} y_t,\quad \sigma_{x_t}^2 \right)
\end{align*}
%
If $\Delta t$ is fixed and $x_t$ is adjusted to follow a Markov chain rather than a Markov process, then this model is equivalent to the CarHMM with normal emission probabilities. Note that all of the parameter transformations above are one-to-one, so it is easy to go from the CarHMM to the continuous model and back again. This allows for the pricipled construction of the continuous-time model to be combined with the computational convienence of the CarHMM.

\subsection{Generalization to unequal time steps}

Using this intuition, we generalize the CarHMM to data with unequal time steps: i.e. $\Delta t$ becomes a function of $t$. First, it is necessary to assume that the behavioral state $x_t$ only changes once between observations, which is a fair assumption if each time step $\Delta t$ is sufficiently small. Next, the following definitions are introduced:
%
$$\Gamma_{\Delta t,\Lambda} = \begin{pmatrix} 
1-e^{-\lambda_1 \Delta t} & p_{12} e^{-\lambda_1 \Delta t} & \dots & p_{1N} e^{-\lambda_1 \Delta t} \\
p_{21} e^{-\lambda_2 \Delta t} & 1-e^{-\lambda_2 \Delta t} & \dots & p_{2N} e^{-\lambda_2 \Delta t} \\
\vdots & \vdots & \ddots & \vdots \\
p_{N1} e^{-\lambda_N \Delta t} & p_{N2} e^{-\lambda_N \Delta t} & \dots  & 1-e^{-\lambda_N \Delta t}
\end{pmatrix}$$
%
where $\Lambda = \{p,\lambda\}$, $p_{ij}$ is the probability that the animal moves from state $i$ to state $j$ given that it left state $i$, and $\lambda_i$ is the rate at which the animal leaves behavioral state $i$. Finally, we just note that the parameters in equation (\ref{CarHMM_to_OU}) become functions of the time step between observations $\Delta t$:
%
\begin{align*}
\mu_{x_t}= \gamma_{x_t}, \qquad \phi_{x_t}(\Delta t) = e^{-\beta_{x_t}\Delta t}, \qquad \sigma^2_{x_t}(\Delta t) = \frac{\omega_{x_t}^2}{2\beta_{x_t}} (1-e^{-2\beta_{x_t}\Delta t})
\end{align*}
%
Then the likelihood of the CarHMM with unequal time steps is just the following:
%
$$\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta) = \delta \prod_{t=2}^{T} \Gamma_{\Delta t,\Lambda} P(y_t;\theta) \mathbf{1}$$
%
where:
%
$$P(y_t;\theta) = \text{diag}(p_\theta(y_t|y_{t-1}, \Delta t, X_t = x_1), . . . , p_\theta(y_t|y_{t-1}, \Delta t, X_t = x_N )), \qquad t > 1$$
%
and:
%
$$p_\theta(y_t|y_{t-1}, \Delta t, X_t = x_t) = \mathcal{N}\left((1-\phi_{x_t}(\Delta t))\mu_{x_t} + \phi_{x_t}(\Delta t) y_{t-1},\quad \sigma_{x_t}^2(\Delta t) \right).$$
%
now $\calL_{\text{CarHMM}}(y;\theta,\Lambda,\delta)$ is simply maximized with respect to $\theta = \{\beta,\omega,\gamma\}$, $\Lambda = \{p,\lambda\}$, and $\delta$ to find the maximum likelihood estimate of the CarHMM generative model with uneven time steps.

### 2020-06-05 Report/Sections/fourier.tex
Many Updates

\section{Capturing Non-Markovian Behavior via the Fourier Transform}

Although the CarHMM and continuous-time model can effectively capture autocorrelation within a process, both still assume Markovian dynamics, i.e. that the observation $Y_t$ depends only the value of the behaviorial state $X_t$ and the previous observation $Y_{t-1}$. However, there are many animal movement process which violate the markov propertey. In particular, on a fine scale, swimming behavior of marine mammals can exhibit periodic behavior as the animal repeatedly flukes to propel itself forward. Work has been done in the past to model non-markovian dynamics in the \textit{behavioral} process \cite{Langrock:2012}, but addressing non-markovian dynamics within observations $Y*$ is still a relatively unstudied area. With improvements in tagging technology allowing for data collection at very high frequencies, noisy and non-markovian fine scale behavior appears to be on the rise.

To address this issue, we recommend borrowing techniques from the signal processing literature to compress the data and summarize its essential elements. In particular, we suggest a method which can capture periodic behavior by decompsing a signal into its fourier components.

Suppose a one-dimensional fine-scale process of length $T^*$, $y^*$ exhibits significant periodic and structured behavior that is clearly non-markovian. It is possible to transform this fine-scale process $y^*$ to a shorter, but higher dimensional process $\hat y^*$ by selecting a window of size $w$, dividing $y^*$ into $\lfloor T^* / w \rfloor$ seperate intervals of length $w$ (truncating $y^*$ appropriately), and taking the discrete fourier transform (DFT) of each interval. The resulting time series $\hat y^*$ will be a complex-valued, $w$-dimensional time series of length $\lfloor T^* / w \rfloor$.

Picking the window length $w$ should be done with care. $w$ should be long enough to capture the periodic behavior of the underlying process (at least twice as long as the length of a period), but short enough so that the resolution of the process remains high and behavioral changes can be captured via the HMM.

A visualization of transforming a one-dimensional sequence $y^*$ to a $w$-dimensional complex sequence $\hat y^*$ can be seen in figure (\ref{fig:fourier_example}).

>>>>

\section{Model Formulation}

\subsection{Short-time Fourier Transform on Fine-scale process}

One issue with HMMs is that they assume markovian dynamics conditioned on the hidden state, i.e. that any observation $Y^*_t$ depends only on the behaviorial state $X^*_t$ (here we focus on the fine-scale process). However, there are many animal movement processes which violate this markov propertey on very fine scales. For example swimming behaviour of marine mammals can be periodic since the animal repeatedly flukes to propel itself forward. Work has been done in the past to model non-markovian dynamics in the \textit{behavioural} process $X^*_t$ \cite{Langrock:2012}, but addressing non-markovian dynamics within the observation process $Y^*_t$ is still a relatively unstudied area. With improvements in tagging technology allowing for data collection at very high frequencies, data exhibiting noisy and non-markovian fine scale behavior is likely to persist in future studies.

To address this issue, we recommend borrowing techniques from the signal processing literature to compress the data and summarize its essential elements. In particular, we suggest performing the discrete-time Short-time Fourier Transform (STFT) over each observed fine-scale process $Y^*_t$:
%
$$STFT\{Y^*_{t,t^*:t^*+w-1}\}(n) := \hat{Y}^{*(n)}_{t,t^*} = \sum_{n = 0}^{w-1} Y^*_{t,t^*+n}e^{-i \frac{2\pi k}{w} n} \qquad \forall n \in \{0, \ldots, w-1\}, \quad t^* \in \{1,w+1,2w+1, \ldots, w \left(\lfloor T^*_t / w \rfloor -1 \right) + 1\}.$$
%
The STFT slides a moving window of length $w$ accross the time series $Y_t^*$ with a step size of $w$ and transforms the domain of each window from time to frequency. This allows the spectrum of $Y_t^*$ at time $t^*$ to be summerized by a $w$-dimensional vector of fourier coeficients. While other step sizes can be used for the sliding window, we select $w$ to avoid serial dependence between windows.

If $Y^*_t \in \mathbb{R}^{T^*_t}$, then $\hat{Y}_t^* \in \mathbb{C}^{\lfloor T^*_t / w \rfloor \times w}$. While this allows $Y^*_t$ to be represneted in a way that eliminates obvious periodic behavior, the dataset itself is still approximately as large as $Y^*_t$ itself. To reduce the size of $\hat{Y}^*_t$, we propose taking summary statistics of each window as follows:
%
$$Z_{t,t^*}^{*(1)} = \mathcal{R}\left(\hat{Y}^{*(0)}_{t,t^*}\right) \qquad Z_{t,t^*}^{*(2)} = \frac{1}{w}\sum_{n=1}^{\tilde{w}}|\hat{Y}^{*(n)}_{t,t^*}|^2$$
%
$Z_{t,t^*}^{*(1)}$ is equal to the average value of $Y_{t,t^*:t^*+w-1}^*$ while $Z_{t,t^*}^{*(2)}$ is equal to the squared 2-norm of $Y_{t,t^*:t^*+w-1}^*$ that can be attributed to frequencies in the signal between $1$ and $\tilde{w}$ periods per window length. Both the window length $w$ and the max frequency $\tilde{w}$ are tuning parameters that should be tuned in a problem-specific way. $w$ should be long enough to capture the periodic behavior of the underlying process (at least as long as the length of a period), but short enough to avoid over-smoothing of the data and to maintain high resultion in the behavioral process $X^*$. $\tilde{w}$ should be selected such that the maximum frequency of $Y_t^*$ that makes biological sense is $\tilde{w}$ per window length. Note that these summary statistics are just one possible choice, and future studies can adjust the definitions as needed. A visualization of transforming a one-dimensional sequence $Y^*$ to $Z^*$ can be seen in figure (\ref{fig:fourier_example}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=7.5in]{../../Plots/fourier_transform.png}
	\caption{Visualization of transforming $Y^*_t$ into $Z^*_t$ using a sliding window and fourier transform.}
	\label{fig:fourier_example}
\end{figure}

Finally, note that it is possible to accomodate for unequal time steps within each window by using the \textbf{non-uniform discrete Fourier transform (NDFT)}. We do not describe the details of this method in this work, but the generalization is straightforward. Refer to Bagchi et al \cite{Bagchi:2001} for details.

\subsection{Model Structure: combining the HHMM and CarHMM}

Hierarchical hidden Markov models can be used to jointly model simultaneous coarse-scale and fine-scale processes taking place simultaneously. However, as mentioned before, the fine-scale process $Y^*$ can often exhibit autocorrelation and intricate structure. Transforming $Y^*_t$ to $Z^*_t$ removes fine-scale periodic behavior, but $Z^*_t$ can still exhibit autocorrelation, especially in the window averge, $Z_{t,t^*}^{*(1)}$. Therefore, we replace the fine-scale HMM within the Hierarchical HMM with a CarHMM according to figure (\ref{fig:CarHHMM}).

\begin{figure}[h!]
	\centering
	\includegraphics[width=7.5in]{../../Plots/CarHHMM.png}
	\caption{Graphical representation of a CarHHMM. The additional arrows representing autocorrelation between observations are shown in red for emphasis.}
	\label{fig:CarHHMM}
\end{figure}

The likelihood of this model is still easy to calculate using the forward algorithm:
%
$$\calL_{\text{CarHHMM}}(y,z^*;\theta,\theta^*,\Gamma,\Gamma^*,\delta,\delta^*) = \delta P(y_1,z_1^*;\theta,\theta^*,\Gamma^*,\delta^*) \prod_{t=2}^T \Gamma P(y_t,z_t^*;\theta,\theta^*,\Gamma^*,\delta^*) \mathbf{1}$$
%
where:
%
\begin{align*}
P(y_t,z_t^*;\theta,\theta^*,\Gamma^*,\delta^*)  = \text{diag}\left[p_{\theta}(y_t|X_t = x_1)\calL_{\text{CarHMM}}\left(z_t^*;\theta^{*(x_1)},\Gamma^{*(x_1)},\delta^{*(x_1)}\right), . . . , \right.\\
\left. p_{\theta}(y_t|x_t = x_N )\calL_{\text{CarHMM}}\left(z_t^*;\theta^{*(x_N)},\Gamma^{*(x_N)},\delta^{*(x_N)}\right) \right]
\end{align*}
%

### 2020-06-05 Report/Sections/simulation.tex
Many Updates

To test this method, dive and acceleration data was simulated using the following procedure:

\begin{enumerate}
	\item 100 dive durations were simulated using a gamma distribution using an HMM generative model and the following parameters:

	\begin{align*}
		\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
		% 
		&Y_t|X_t \sim \text{Gamma} \\
		\bbE(Y_t|X_t = 1) &= 15, \quad \bbE(Y_t|X_t = 2) = 60 \\
		%
		\bbV(Y_t|X_t = 1) &= 25, \quad \bbV(Y_t|X_t = 2) = 100
	\end{align*}
\end{enumerate}

Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive. The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly. As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval. Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval. Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken. To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored. One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation. The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find. An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.

\subsection{Lag Plot}

In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum. The results of doing so are shown in figure (\ref{fig:lag}). Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation. While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.

>>>>

\subsection{Data Simulation}

To test the CarHHMM with STFT, a sequence of 100 marine-animal dives were simulated. The coarse-scale observations $Y$ were set as the duration of each dive, and the fine-scale observations $Y^*$ were set as one dimensional acceleration readings simulated at 50 hertz. Specifically, the following procedure was followed:

\begin{enumerate}
	\item 100 dive durations were simulated using an HMM generative model with the following parameters:

	\begin{align*}
		\Gamma &= \begin{pmatrix} 0.4 & 0.6 \\ 0.6 & 0.4 \end{pmatrix} \\
		% 
		&Y_t|X_t \sim Gamma \\
		\bbE(Y_t|X_t = 1) &= 15 s, \quad \bbE(Y_t|X_t = 2) = 60 s \\
		%
		\bbV(Y_t|X_t = 1) &= 25 s^2, \quad \bbV(Y_t|X_t = 2) = 100 s^2
	\end{align*}

	\item Once the dive durations were calculated, for each $t \in \{1, \ldots, 100\}$, dive $t$ was broken into $\lfloor Y_t/2 \rfloor$ 2-second segments (the end of the dive sequence was discarded). Further, each 2-second segment was assigned a behaviour according to a fine-scale Markov chain $X^*_t$, where $X^*_{t,t^*} \in \{1,2\}$ and $t^* \in \{1,101,201,\ldots,100*\lfloor Y_t/2 \rfloor + 1\}$. The parameters of the fine-scale behaviour Makov chain were set to be as follows:
	%
	$$\Gamma^{*(1)} = \begin{pmatrix} 0.25 & 0.75 \\ 0.75 & 025 \end{pmatrix} \qquad 	\Gamma^{*(2)} = \begin{pmatrix} 0.75 & 0.25 \\ 0.25 & 0.75 \end{pmatrix}$$
	%
	where $\Gamma^{*(1)}$ was used for dives where $X_t = 1$ and $\Gamma^{*(2)}$ was used for dives where $X_t = 2$

	\item For each 2-second segment, the Fourier modes $\hat{Y}^*_{t,t^*}$ were simulated using the following procedure. Note that the $n^{th}$ Fourier mode of $\hat{Y}^*_{t,t^*}$ is denoted as $\hat{Y}^{*(n)}_{t,t^*}$ :

	\begin{align*}
	\hat{Y}^{*(0)}_{t,1} &\sim \mathcal{N} \left(\mu = 1, \sigma^2 = 0.01 \right) & \\
	%
	\hat{Y}^{*(0)}_{t,t^*} &\sim \mathcal{N} \left(\mu = 0.9 Y^{*(0)}_{t,t^*-100} + 0.1, \sigma^2 = 0.01 \right), & t^* \in \{101,201,\ldots, 100*\lfloor Y_t/2 \rfloor + 1\} \\
	%
	\hat{Y}^{*(n)}_{t,t^*} &= a_{t,t^*}^{(n)} i\sqrt{b^{(n)}_{t,t^*}}, & n \in \{1,\ldots,49\} \\\\
	%
	%
	a_{t,t^*}^{(n)} &\sim  \left\{\begin{array}{lr}
	-1 & w.p. \enspace 1/2 \\
	1  & w.p. \enspace 1/2
	\end{array}\right. \\
	%
	(b^{(n)}_{t,t^*}|X^*_{t,t^*}  = 1) &\sim Gamma(1/n^2, 1) & \\
	%
	(b^{(n)}_{t,t^*}|X^*_{t,t^*} = 2) &\sim  \left\{\begin{array}{lr}
	Gamma(1/n^2, 1) & \text{for } n \neq 2\\
	Gamma(100,1) & \text{for } n = 2
	\end{array}\right. & \\\\
	%
	%
	\hat{Y}^{*(50)}_{t,t^*} &= 0 & \\
	%
	\hat{Y}^{*(n)}_{t,t^*}  &= -\hat{Y}^{*(100-n)}_{t,t^*}, & \qquad n \in \{51,\ldots,99\}
	\end{align*}

	Finally, $Y^*_{t,t^*:t^*+99}$ is set using the inverse discrete fourier transform of $\hat{Y}^*_{t,t^*}$:

	$$Y^*_{t,t^*:t^*+99} = IDFT\left(\hat{Y}^*_{t,t^*}\right)$$

	This gives a strong periodic component to acceleration when the subdive state $X^*_{t,t^*}= 2$. There are several practical reasons behind this construction:

	\begin{enumerate}
		\item $\hat{Y}^*_{t,t^*}$ is anti-symetric about $\hat{Y}^{*(50)}_{t,t^*}$ so that its inverse fourier transform is real-valued.
		\item $\hat{Y}^{*(n)}_{t,t^*}$ decays like $1/n^2$ to facilitate continutity.
	\end{enumerate}

\end{enumerate}

Note that this process idoes not result in a continuous sequence $Y^*_t$ since the average values of consecutive 2-second segments jump. However, note that these average values are highly autocorrelated, so the jumps are not too severe. See figure (\ref{fig:sim_data}) for details. In addition, $Y^*_t$ has the following desirable properties:

\begin{enumerate}
	\item $Z^{*(1)}_{t,t^*+1} | Z^{*(1)}_{t,t^*} \sim \mathcal{N} \left(\mu = 0.9 Z^{*(1)}_{t,t^*} + 0.1, \sigma^2 = 0.01 \right)$
	\item $$Z^{*(2)}_{t,t^*} \sim \left\{\begin{array}{lr} 
	\text{Gamma}(\sum_{n=1}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 1 \\
	\text{Gamma}(20 + \sum_{n=2}^{\tilde{w}} \frac{1}{n^2},1) & \text{for } X^*_{t,t^*} = 2
	\end{array}\right. $$
\end{enumerate}

So we can directly compare the results of the CarHHMM with the ground truth.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=5in]{../../Plots/sim_data.png}
	\caption{Simulated Acceleration Data.}
	\label{fig:sim_data}
\end{figure}
I still need to put in the actually results from the simulation study- I had to rethink how to do the simulation in the first place since enforcing continuity was REALLY hard to do in a principled way.

\iffalse
\subsection{Model Misspecification}

There are two ways to simulate the fourier modes of the acceleration vector $\bfA_{s,t}$:

\begin{enumerate}
	\item $$\bbE_1\left(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}\right) = \phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} $$

	\item $$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
	\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
	\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
	\end{array}\right\}$$
\end{enumerate}

\subsubsection{$\bbE_1$}

First, look at $\bbE_1$. This comes with an issue in that when $X^*_{s,t}$ changes values, it will take a while for the process to ``converge" long-run mean $\mu_{X^*_{s,t}}$. In particular, if the behavioral process $X^*$ had been in state 1 for a while and then switches to state 2 for a while, then:

\begin{align*}
	\bbE_1\left(\bfA_{s+1,t}|X^*_{s:s+1,t} = 2, X^*_{s,t} = 1\right) &=  \phi \mu_1 + (1-\phi) \mu_2 \\
	%
	\bbE_1\left(\bfA_{s+2,t}|X^*_{s:s+2,t} = 2, X^*_{s,t} = 1\right) &= \phi\left( \phi \mu_1 + (1-\phi) \mu_2\right) + (1-\phi)\mu_2 \\
	&= \phi^2 \mu_1 + (1-\phi^2) \mu_2\\
	%
	\bbE_1\left(\bfA_{s+n,t}|X^*_{s:s+n,t} = 2, X^*_{s,t} = 1\right) &=  \phi^n \mu_1 + (1-\phi^n) \mu_2 \\
\end{align*}

So, if using $\bbE_1$, there is an assumption that the fourier modes of the acceleration vector $\bfA$ linearly ``converge" to the new value $\mu_i$ from $\mu_j$ when $X^*_{s,t}$ changes from $i$ to $j$. An example of this can be seen in figure (\ref{fig:FOVEDBA}).

 \begin{figure}[h!]
 	\centering
 	\includegraphics[height=3in]{../Code/test.png}
 	\caption{The FoVeDBA of simulated data using $\bbE_1$. Note that the FoVeDBA approaches $\approx$ 300 in state 2 from $\approx$ 10 in state 1 with roughly linear convergence.}
 	\label{fig:FOVEDBA}
 \end{figure}

 Note that if the generative model is correlated, but the inference model isn't, then quick jumps between states 1 and 2 can be missed because the chain never ``converges" to the new mean $\mu_i$, as shown in figure (\ref{fig:STATE}).

\begin{figure}[h!]
	\centering
	\includegraphics[height=3in]{../Code/test2.png}
	\caption{The probability of being in each behavioral state $X^*_{s,t}$ as estimated by the HHMM. The colors represent the true behavioral state.}
	\label{fig:STATE}
\end{figure}

 \subsubsection{$\bbE_2$}

Clearly, the model has issues decodeing the behavioral state when the model is misspecified. Therefore, if the true generative process does not have correlation when jumping between states, then it is better to use $\bbE_2$:

$$\bbE_2(\bfA_{s,t} | \bfA_{s-1,t}, X^*_{s,t}, X^*_{s-1,t}) = \left\{\begin{array}{lr}
\phi \bfA_{s-1,t} + (1-\phi) \mu_{X^*_{s,t}} & \text{for } X^*_{s,t} = X^*_{s-1,t}\\
\mu_{X^*_{s,t}} & \text{for } X^*_{s,t} \neq X^*_{s-1,t}
\end{array}\right\}$$

This fixes the issue of having to ``converge" to the new value of $\mu_i$ when jumping to state $i$. However, note that $\bbE_2$ depends upon both $X^*_{s,t}$ and $X^*_{s-1,t}$. This breaks the assumption in the forward algorithm that $\bfA_{s,t}$ depends only on $X^*_{s,t}$. One way to deal with this issues is to introduce a new hidden state $Z^*_{s,t} = (X^*_{s-1,t},X^*_{s,t})$. This fixes the issue, but now $Z^*_{s,t}$ can take one of $s^2$ values instead of $s$. This grows the state space of our model and can make inference harder.

 \fi

%Figure (\ref{fig:smoothing}) also shows the calculated ODBA as a function of time for one a specific dive. The acceleration exhibits sinusoidal behavior at several points in time which cannot be modeled using HMMs straightforwardly. As a result, the data was split into two-second intervals (100 data points each), and each interval was summarized by the Fourier transform of the ODBA and the velocity at the \textit{end} of the time interval. Velocity was taken at the end of the time interval because the behavior of the whale during the time interval has no impact on the velocity of the whale at the beginning of the time interval. Because velocity is highly correlated with the average acceleration within an interval, the mean was subtracted out of each ODBA time series before the Fourier transform was taken. To reduce the dimension of the ODBA Fourier transform, Fourier coefficients corresponding to frequencies less than or equal to 5 hertz were summed and frequencies above 5 hertz were ignored. One nice property of this process is that it reduces the number of data points by a factor of 100, which drastically speeds up parameter estimation. The optimal length of the time interval over which to take the Fourier transform is a tuning parameter that is difficult to find. An ideal time interval should be long enough give useful Fourier coefficients, but short enough to preserve as much information about the vertical velocity of the animal as possible.

%\subsection{Lag Plot}

%In order to find the appropriate HMM to model whale behavior, a lag plot was made for both velocity and the ODBA fourier sum. The results of doing so are shown in figure (\ref{fig:lag}). Unfortunately, the number of behavioral states is not clear from the lag plot, but it is clear that the velocity exhibits a large degree of autocorrelation. While the ODBA Fourier sum also exhibits some autocorrelation, the relationship is less strong, so autocorrelation was not incorporated in the ODBA Fourier sums emission distribution.
